Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-01 paper/metamonads.tex
some re-writing

let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
\end{code}
\caption{Our monad}
\end{figure*}

@@

\oleg{Should we rename lif into ifL?}
Now we can build more complex monadic operations.
\begin{code}
let lif test th el = ret .< if .~test then .~th else .~el >.
\end{code}

ifM and ifL: show example. Mention 'ifM' and `reset'. Investigating
the connection with the delimited continuations: for future work.

What is state and why we need it: making aspects interleave: use det
as an example.

@@

Making state modular and composable: several aspects can contribute
to the state, without interference (show: Rank, Det, Permutation
aspects).

>>>>

let l1 f = fun x -> mdo { t <-- x; f t}
let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}

let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.

let lif test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k ->
  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}
\end{figure*}

@@

\oleg{Should we rename lif into ifL?}
We can write operations that generate code other than let-statements,
e.g., conditionals: |lif| (Figure XXX). The function |lif|, albeit
straightforward, is not as general as we wish: its argument are
already generated pieces of code rather than monadic values. So, we
may want to ``lift it'', see |ifM'|:

\begin{code}
let ifM' test th el = mdo {
  testc <-- test;
  thc   <-- th;
  elc   <-- el;
  lif testc thc elc}
\end{code}

\omitnow{
\begin{code}
let ifM' test th el = bind test (fun testc ->
                      bind th   (fun thc ->
                      bind el   (fun elc ->
		      lif testc thc elc)));;
\end{code}
}

We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
We soon notice the need in another |ifM| function, with tha same
interface (see ifM in Fig. XXX). The difference between them is
apparent from the following:

\begin{code}
let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
                           (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
;;
.<fun a_1 ->
   fun i_2 ->
    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
\end{code}

whereas if we replace |ifM'| with |ifM|, the result will be

\omitnow{
\begin{code}
let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
                           (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
;;
\end{code}
}
\begin{code}
.<fun a_1 ->
   fun i_2 ->
    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
\end{code}

The difference is apparent: in the code with |ifM'|, the let-insertion
happend \emph{before} the if-expression, that is, before the test that
the index |i| is positive. That is, if |i| turned out
negative, |a.(i)| will generate the out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects, when generating control constrcutions such as conditionals.
The form |ifM| handles the such effects correctly. In our code, we
need similar operators for other Ocaml control forms: for generating
case-matching statements and for- and while-loops. To make writing
such code general and independent of the implementation of the monad,
we define a function XXX. As the name implies, we do use control
effect implemented via essentially delimited continuations (cite?).

@@

Making state modular and composable: several aspects can contribute
to the state, without interference (show: Rank, Det, Permutation
aspects).
What is state and why we need it: making aspects interleave: use det
as an example. State as an open record. List of polymorphic variants
appeared to be the easiest way to implement such a union, in a pure
functional way.
